---
title: "Workshop2_KeystoneExercise"
author: "Hannah Brown"
format: html
editor: visual
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

# Module 2 - Exercise 2.12

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

# Introduction

### Here we are going to wrangle in messy data on the Ross River estuary fish surveys and produce final clean tables and figure. We have 4 separate files that were provided to us by a former PhD student. We know we have data for 4 sites, 4 species, and 30 days of sampling. We also have metadata on the sites and sensor data on the environmental conditions at each site.

## Housekeeping

```{r}
rm(list=ls())
```

## Packages

```{r}
library(tidyverse)
library(readxl)
```

# Load files

## Load Metadata

```{r}
metadata <- read.csv("../data/workshop2/estuary_metadata.csv")
```

### Inspect metadata

```{r}
head(metadata)
glimpse(metadata)
```

## Load Main data set

```{r}
catch1 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet =1)
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet =2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet =3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet =4)

```

### Bind rows together

```{r}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)

#Clean up environement
rm(catch1, catch2, catch3, catch4)
```

### Examine data

```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

### Find site and species variables in data

```{r}
catch_data |>
  distinct(site)

catch_data |>
  distinct(species)
```

### Clean up site name in catch and meta data to same variables

```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  ) 

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```

### Recheck name

```{r}
catch_data |> 
  str()
catch_data |> 
  distinct(site)
glimpse(metadata)
```

## Load taxonomic dictionary

```{r}
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
head(species_dict)
```

### Clean up species name in catch data to match taxonomic dictionary

```{r}
catch_data <- catch_data |>
  mutate(
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```

### Check common name matches in catch data and taxonomic dictionary

```{r}
catch_data |> 
  distinct(species)
species_dict |>
  distinct(common_name)
```

### Taxonomic translation (common to species)

```{r}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name))
```

### Check species names in catch data

```{r}
glimpse(catch_data)
```

## Load sensor data

```{r}
sensor_raw <- read.csv("../data/workshop2/estuary_sonde_data.csv")
```

### Check sensor data

```{r}
head(sensor_raw)
str(sensor_raw)
glimpse(sensor_raw)
summary(sensor_raw)
```

### Clean up sensor data

```{r}
sensor_data <- sensor_raw |>
  mutate(
    # covert timestamp to date time
    datetime = dmy_hm(timestamp),
    # convert the -999 turbidity (error) values to NA
    turbidity = na_if(turbidity, -999),
    # convert negative salinity values to NA
    salinity = ifelse(salinity < 0, NA, salinity)
  )
```

### Summarise 15-minute sonde data into daily means

```{r}
sensor_daily <- 
  sensor_data |>
  mutate(date = as_date(floor_date(datetime, "day"))) |>
  group_by(site, date) |>
  summarise(
    mean_temp = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    mean_turbidity = mean(turbidity, na.rm = TRUE),
    .groups = "drop")
```

## Join catch data with metadata and sensor data

```{r}
catch_data_final <- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_daily, by = join_by(site, date))
```

## Wrangle final data set

```{r}
catch_data_final |> 
  complete(
    nesting(site, date, 
           # lat, lon, 
            zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name
  ) |>
  mutate(
    # Convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )
```

# Final data set is now complete.

# Lets now look at the difference in environmental conditions across these zones.

## Salinity

```{r}
# Let's make a table
sal_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
   # sem_salinity = sd(mean_salinity) / sqrt(n()),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

# And from this a plot
sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
```

### There is a clear salinity gradient across the zones, with the upper estuary being fresher and the lower estuary being more saline. This is consistent with what we would expect in an estuarine system, where freshwater inputs from rivers mix with seawater from the ocean. This salinity gradient could be an important driver of fish abundance patterns across these zones, as different species may have different salinity preferences or tolerances.

## Turbidity

```{r}
turb_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_turbidity = sd(mean_turbidity, na.rm = TRUE),
    median_turbidity = median(mean_turbidity, na.rm = TRUE),
    .groups = "drop"
  )

turb_sum |>
  ggplot(aes(x = zone, y = median_turbidity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_turbidity - sd_turbidity, ymax = median_turbidity + sd_turbidity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Turbidity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Turbidity (NTU)")
```

### There is minimal tubidity differences across the zones, with all zones having similar median turbidity values. This suggests that turbidity may not be a major driver of fish abundance patterns across these zones, although it could still have some influence on certain species that are more sensitive to turbidity levels.

## Temperature

```{r}
temp_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_temperature = sd(mean_temp, na.rm = TRUE),
    median_temperature = median(mean_temp, na.rm = TRUE),
    .groups = "drop"
  )

temp_sum |>
  ggplot(aes(x = zone, y = median_temperature)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_temperature - sd_temperature, ymax = median_temperature + sd_temperature), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Temperature by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Temperature (°C)")

```

### There is a slight temperature gradient across the zones, with the upper estuary being slightly warmer than the lower estuary. However, the differences in median temperature across the zones are relatively small. This suggests that temperature may not be a strong driver of fish abundance patterns across these zones.

## Salinity is the most pronounced environmental gradient across these zones, and is likely to be the most important driver of fish abundance patterns in this estuarine system. 

## Remake salinity plot presentable for publication

```{r}
#| fig-cap: "Salinity Gradient across zones. Ross River Estuary System, Townsville, NQ, Australia May 2026."

sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
  
```

## We can now turn to looking at how fish abundance varies across these zones, and whether it is related to the salinity gradient we have observed.

### Calculate summary stats

```{r}
fish_stats <- catch_data_final |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count),
    sem_catch = sd(count) / sqrt(n()),
    .groups = "drop"
  )
```

### Plot fish abundance across zones

```{r}
fish_stats |> 
  ggplot(aes(x = zone, y = mean_catch, fill = zone)) +
  geom_line(group = 1) +
  # add in points from master data
  geom_point(data = catch_data_final, aes(x = zone, y = count), 
             position = position_jitter(width = 0.2), alpha = 0.2, size = 1, inherit.aes = FALSE) +
  geom_point(position = "dodge", size = 4, pch = 23) +
  
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), 
                position = position_dodge(0.9), width = 0.2) +
  
  facet_wrap(~ scientific_name, scales = "fixed") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey20"),
   # legend.position = "bottom"
    legend.position = "none"
  ) +
  labs(
    #title = "Predatory Fish Abundance across Estuarine Salinity Gradients",
    #subtitle = "Ross River Estuary System, May 2026",
    x = "Spatial Zone",
    y = "Abundance (per day)",
    fill = "Spatial Zone"
  )
```
