---
title: "Workshop2_KeystoneExercise"
author: "Hannah Brown"
format: html
editor: visual
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

## Module 2 - Exercise 2.12

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  warning = FALSE)
```

#Introduction Here we are going to wrangle messy data on the Ross River estuary fish surveys and produce final clean tables and figures

# Housekeeping

```{r}
rm(list=ls())
```

# Packages

```{r}
library(tidyverse)
library(readxl)
```

# Load files

## Metadata

```{r}
metadata <- read.csv("../data/workshop2/estuary_metadata.csv")
```

# Inspect metadata

```{r}
head(metadata)
glimpse(metadata)
```

## Main dataset

```{r}
catch1 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet =1)
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet =2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet =3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet =4)

```

# Bind rows together

```{r}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)

#Clean up environement
rm(catch1, catch2, catch3, catch4)
```

# Examine data

```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

# Find site and species variables in data

```{r}
catch_data |>
  distinct(site)

catch_data |>
  distinct(species)
```

# Clean up site name

```{r}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  ) 

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```

# Recheck name

```{r}
catch_data |> 
  str()
catch_data |> 
  distinct(site)
glimpse(metadata)
```

# Load taxonomic dictionary

```{r}
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
head(species_dict)
```

# Load sensor data

```{r}
sensor_raw <- read.csv("../data/workshop2/estuary_sonde_data.csv")
```

# Check sensor data

```{r}
head(sensor_raw)
str(sensor_raw)
glimpse(sensor_raw)
summary(sensor_raw)
```

# Clean up sensor data

```{r}
sensor_data <- sensor_raw |>
  mutate(
    # covert timestamp to date time
    datetime = dmy_hm(timestamp),
    # convert the -999 turbidity (error) values to NA
    turbidity = na_if(turbidity, -999),
    # convert negative salinity values to NA
    salinity = ifelse(salinity < 0, NA, salinity)
  )
```

# Summarise 15-minute sonde data into daily means

```{r}
sensor_daily <- 
  sensor_data |>
  mutate(date = as_date(floor_date(datetime, "day"))) |>
  group_by(site, date) |>
  summarise(
    mean_temp = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity, na.rm = TRUE),
    mean_turbidity = mean(turbidity, na.rm = TRUE),
    .groups = "drop")
```

# Clean up species name in catch data to match taxonomic dictionary

```{r}
catch_data <- catch_data |>
  mutate(
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```

# Check common name matches in catch data and taxonomic dictionary

```{r}
catch_data |> 
  distinct(species)
species_dict |>
  distinct(common_name)
```

# Taxonomic translation (common to species)

```{r}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name))
```

# Join catch data with metadata and sensor data

```{r}
catch_data_final <- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_daily, by = join_by(site, date))
```

```{r}
catch_data_final |> 
  complete(
    nesting(site, date, 
           # lat, lon, 
            zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name
  ) |>
  mutate(
    # Convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )
```

# Check for salinity differences between zones

```{r}
sal_sum <-
catch_data_final |> 
  group_by(scientific_name, zone) |>
  summarise(
    sem_salinity = sd(mean_salinity, na.rm = TRUE) / sqrt(n()),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    )

# or with plotrix
library(plotrix)
catch_data_final |>
  group_by(scientific_name, zone) |>
  summarise(
    sem_salinity = std.error(count),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )
```

# Make point and error bar plot of salinity by zone for each species

```{r}
sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
  
```

# Calculate fish catch abundance statistics (mean and standard error) by zone and species

```{r}
fish_stats <- catch_data_final |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count),
    sem_catch = sd(count) / sqrt(n()),
    .groups = "drop"
  )
```

# Plot fish catch abundance by zone and species with standard error

```{r}
fish_stats |> 
  ggplot(aes(x = zone, y = mean_catch, fill = zone)) +
  geom_point(position = "dodge", size = 2, pch = 23) +
  #geom_line(group = 1) +
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), 
                position = position_dodge(0.9), width = 0.2) +
  # add in points from master data
  # geom_point(data = catch_data_final, aes(x = zone, y = count), 
  #           position = position_jitter(width = 0.2), alpha = 0.5, size = 1, inherit.aes = FALSE) +
  facet_wrap(~ scientific_name, scales = "fixed") +
  theme_bw() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey20"),
   # legend.position = "bottom"
    legend.position = "none"
  ) +
  labs(
    #title = "Predatory Fish Abundance across Estuarine Salinity Gradients",
    #subtitle = "Ross River Estuary System, May 2026",
    x = "Spatial Zone",
    y = "Abundance (per day)",
    fill = "Spatial Zone"
  )
```
